---
title: "Exploring Car Performance and Fuel Efficiency"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring Car Performance and Fuel Efficiency}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(assignment4shinykkapoor)
library(ggplot2)
library(dplyr)
```

# Introduction

This vignette demonstrates how to explore the cleaned cars_data dataset
included in the assignment4shinykkapoor package. The dataset extends the
classic mtcars data with new variables such as km_per_litre (fuel
efficiency) and power_to_weight.

We aim to communicate the relationship between engine power and fuel
efficiency.

# The Data

```{r}
head(get_cars_data())
```

## Key variables

-   hp: engine horsepower (higher = more powerful engine)

-   km_per_litre: fuel efficiency in kilometres per litre (higher =
    cheaper to run)

-   am: transmission type (“Automatic” or “Manual”)

-   power_to_weight: horsepower divided by weight (how strong the engine
    is relative to how heavy the car is)

# Relationship Between Power and Efficiency

```{r}
get_cars_data() |>
  ggplot(aes(x = hp, y = km_per_litre, color = am)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    x = "Horsepower",
    y = "Fuel efficiency (km per litre)",
    color = "Transmission",
    title = "Higher horsepower usually lowers fuel efficiency"
  ) +
  theme_minimal()
```

## Interpretation

We see a clear negative relationship: as horsepower increases, fuel
efficiency (km per litre) tends to drop.

-   Manual cars (often lighter) appear slightly more efficient at
    comparable horsepower levels.

## Interactive Exploration 

You can explore these same relationships interactively using the
dashboard bundled with this package:

```         
library(assignment4shinykkapoor)
run_app()
```

The Shiny app allows you to:

-   filter cars by cylinder count and horsepower range,

-   visually compare horsepower vs fuel efficiency

-   inspect a sortable data table of matching cars,

-   read an English summary of the filtered subset.

This makes the analysis accessible to someone who doesn’t use R, which
is the core goal of ETC5523: communicating data clearly, not just
calculating it.
